BayesPy: Variational Bayesian Inference in Python

### 2015-03-11 bayespy.tex
Add BayesPy vs Infer.NET comparison

BayesPy can be used to construct a wide range of conjugate exponential family
models.  The documentation provides detailed examples of how to construct, for
instance, regression models, principal component analysis models, linear
state-space models, mixture models and hidden Markov models.  In addition,
BayesPy has been used in two publications about parameter expansion and
time-varying dynamics for linear state-space models
\citep{Luttinen:2013,Luttinen:2014}, which show that it is applicable to large
complex models requiring efficient inference.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node [obs] (yn) {$y_n$};
    \node [latent, above left=of yn] (mu) {$\mu$};
    \node [latent, above right=of yn] (tau) {$\tau$};

    \edge {mu, tau} {yn};

    \tikzstyle{plate caption} = [caption, node distance=0, inner sep=0pt,
    below=5pt of #1.south]

    \plate {} {(yn)} {$n=1,\ldots,10$};
  \end{tikzpicture}
  \caption{Graphical model of the example problem.}
  \label{fig:graphical_model}
\end{figure}
In order to demonstrate BayesPy, this section solves an extremely simple problem
but which includes the main steps of using BayesPy.  The task is to estimate the
unknown mean and precision parameters of a Gaussian distribution given ten
observations.  Thus, the likelihood function is Gaussian:
\begin{align*}
  p(y_1,\ldots,y_{10}|\mu,\tau) &= \prod^{10}_{n=1}\mathcal{N}(y_n | \mu, \tau),
\end{align*}
where the Gaussian distribution is parameterized by precision instead of
variance.  The unknown mean $\mu$ and precision $\tau$ are given broad Gaussian
and gamma priors, respectively:
\begin{align*}
  p(\mu) &= \mathcal{N}(\mu|0, 10^{-6}), & p(\tau) &= \mathcal{G}(\tau|10^{-6},
  10^{-6}).
\end{align*}
Figure~\ref{fig:graphical_model} shows the graphical model of this simple model.
The model is constructed in BayesPy by creating nodes:
\begin{lstlisting}
from bayespy.nodes import GaussianARD, Gamma
mu = GaussianARD(0, 1e-6)
tau = Gamma(1e-6, 1e-6)
\end{lstlisting}
These two nodes represent the two unknown variables that we are interested in.
They are the parameters of the Gaussian distribution of the observed variable:
\begin{lstlisting}
y = GaussianARD(mu, tau, plates=(10,))
\end{lstlisting}
The number of observations is given as plates which define the number of
repetitions.  Note that the syntax of the model construction is similar to the
mathematical formulation which defines the conditional distributions for each
variable.  Thus, it is easy to understand the model from the Python code.  After
the model has been constructed, some nodes are marked as observed and given
data:
\begin{lstlisting}
y.observe([4.5, 3.9, 6.3, 5.6, 4.9, 2.8, 7.4, 6.1, 4.8, 2.1])
\end{lstlisting}
Next, we want to find the posterior approximation for our latent variables.  We
create the VB inference engine:
\begin{lstlisting}
from bayespy.inference import VB
Q = VB(y, mu, tau)
\end{lstlisting}
The inference engine takes all the stochastic nodes of the model as input.  The
message passing algorithm can be run by calling \texttt{update} method:
\begin{lstlisting}
Q.update(repeat=100)
\end{lstlisting}
The algorithm is run for 100 iterations or until it converges.  In this case,
the algorithm converges in four iterations.  The posterior approximation can be
examined, for instance, by plotting the probability density functions of the
latent variables:
\begin{lstlisting}
import bayespy.plot as bpplt
import numpy as np
bpplt.pyplot.figure()
bpplt.pdf(Q[mu], np.linspace(0, 10, num=300))
bpplt.pyplot.figure()
bpplt.pdf(Q[tau], np.linspace(0, 2, num=300))
bpplt.pyplot.show()
\end{lstlisting}
Figure~\ref{fig:posterior} shows the resulting approximate posterior
distributions.  The module \texttt{bayespy.plot} contains functions for simple
plotting tasks, such as plotting the probability density function or the Hinton
diagram of a variable.  Note that the standard \texttt{pyplot} module of
Matplotlib is available at the \texttt{plot} module of BayesPy.
\begin{figure}
  \begin{center}
    \begin{subfigure}[b]{0.4\textwidth}
      \includegraphics[width=\textwidth]{fig_mu}
      \caption{$q(\mu)$}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
      \includegraphics[width=\textwidth]{fig_tau}
      \caption{$q(\tau)$}
    \end{subfigure}
  \end{center}
  \caption{The posterior approximations of $\mu$ and $\tau$.}
  \label{fig:posterior}
\end{figure}

>>>>

BayesPy can be used to construct conjugate exponential family models.  The
documentation provides detailed examples of how to construct, for instance,
regression models, principal component analysis models, linear state-space
models, mixture models and hidden Markov models.  In addition, BayesPy has been
used in two publications about parameter expansion and time-varying dynamics for
linear state-space models \citep{Luttinen:2013,Luttinen:2014}, which show that
it is applicable to large complex models requiring efficient inference.

@@

This section demonstrates BayesPy with a Gaussian mixture model.  A simple
artificial dataset is created by drawing 200 samples from a 2-dimensional
Gaussian distribution.  80 samples have mean $[1, 1]$ and 120 samples have mean
$[0, 0]$:
\begin{lstlisting}
import numpy as np
N = 200, D = 2
data = np.random.randn(N, D)
data[:80,:] += np.ones(D)
\end{lstlisting}
We construct a mixture model for the data and assume that the parameters and the
cluster assignments are unknown, even the true number of clusters.  The model
uses a maximum number of five clusters but the effective number of clusters will
be determined automatically:
\begin{lstlisting}
K = 5
\end{lstlisting}
The unknown cluster means and precision matrices are given Gaussian and Wishart
prior distributions:
\begin{lstlisting}
from bayespy import nodes
mu = nodes.Gaussian(np.zeros(D), 0.01*np.identity(D), plates=(K,))
Lambda = nodes.Wishart(D, 0.01*np.identity(D), plates=(K,))
\end{lstlisting}
The \texttt{plates} keyword argument is used to define repetitions similarly as
the plate notation in graphical models.  The cluster assignments are categorical
variables, and the cluster probabilities are given a Dirichlet prior
distribution:
\begin{lstlisting}
alpha = nodes.Dirichlet(0.01*np.ones(K))
z = nodes.Categorical(alpha, plates=(N,))
\end{lstlisting}
The observations are from a Gaussian mixture distribution:
\begin{lstlisting}
y = nodes.Mixture(Z, nodes.Gaussian, mu, Lambda)
\end{lstlisting}
The second argument for the \texttt{Mixture} node defines the type of the
mixture distribution, in this case Gaussian.  The variable is marked as observed
by providing the data:
\begin{lstlisting}
y.observe(data)
\end{lstlisting}
Next, we want to find the posterior approximation for our latent variables.  We
create the variational Bayesian inference engine:
\begin{lstlisting}
from bayespy.inference import VB
Q = VB(y, mu, z, Lambda, alpha)
\end{lstlisting}
Before running the VMP algorithm, the symmetry in the model is broken by
a random initialization of the cluster assignments:
\begin{lstlisting}
z.initialize_from_random()
\end{lstlisting}
Without the random initialization, the clusters would not be separated.  The VMP
algorithm is run for 100 iterations or until convergence:
\begin{lstlisting}
Q.update(repeat=100)
\end{lstlisting}
The results can be examined, for instance, by printing the posterior
distributions of some variables:
\begin{lstlisting}
print(alpha)
print(mu)
\end{lstlisting}
In addition, module \texttt{bayespy.plot} contains several plotting functions to
examine the results visually.

\section{Comparison with Infer.NET}

This section provides a brief comparison with Infer.NET because it is another
active project implementing the variational message passing algorithm, whereas
the other related active projects implement MCMC.  The main advantages of
Infer.NET over BayesPy are its support for non-conjugate models and other
inference engines.  On the other hand, BayesPy is open-source software and
supports several advanced VB methods (e.g., stochastic variational inference and
collapsed inference).

The speed of the packages were compared by using two widely used models: a
Gaussian mixture model (GMM) and principal component analysis
(PCA).\footnote{The scripts for running the experiments are available in the
  supplementary material.}  Both models were run for small and large
artificially generated datasets.  For GMM, the small model used 10 clusters for
200 observations with 2 dimensions, and the large model used 40 clusters for
2000 observations with 10 dimensions.  For PCA, the small model used
10-dimensional latent space for 500 observations with 20 dimensions, and the
large model used 40-dimensional latent space for 2000 observations with 100
dimensions.  Infer.NET used a fully factorizing approximation for the PCA model,
whereas BayesPy did not factorize with respect to the latent space dimensions.
The experiments were run on a Linux computer with a quad-core processor
(i7-4702MQ).  Because the packages run practically identical algorithms and thus
converged to similar solutions in a similar number of iterations (50--100
iterations depending on the dataset), we compare the average CPU time per
iteration.

The results are summarized in Table~\ref{tab:speed}.  On the smaller datasets,
Infer.NET is slightly faster because BayesPy has some overhead from using
Python.  However, on the larger datasets, BayesPy is significantly faster
because it uses BLAS and LAPACK libraries for fast numerical
operations.\footnote{NumPy must use BLAS/LAPACK because BayesPy uses NumPy for
  numerical operations.}  In addition, if using multi-threaded BLAS/LAPACK, the
difference in wall-clock time can be even larger than in CPU time.  The speed of
BayesPy for these models could potentially be improved by using collapsed or
stochastic variational inference, and the accuracy could be improved by using
Gaussian-Wishart nodes which do not factorize with respect to the Gaussian and
Wishart variables.

\begin{table}[tb]
  \centering
  \caption{The average CPU time (in milliseconds) per iteration for each dataset.}
  \small
  \begin{tabular}{ccccc}
    &
    Small GMM
    &
    Large GMM
    &
    Small PCA
    &
    Large PCA
    \\
    \hline
    BayesPy     & 45 & 210  & 52 & 380
    \\
    Infer.NET   & 26 & 4400 & 37 & 2200
  \end{tabular}
  \label{tab:speed}
\end{table}

% related active software projects discussed in the introduction,

% Because the other
% active Bayesian software packages discussed in the introduction do not implement
% the variational message passing algorithm,

%* compiled

%* non-conjugate

%* other engines

%* open source

%* advanced

%* dynamic nodes

% \begin{figure}
%   \centering
%   \begin{tikzpicture}
%     \node [obs] (yn) {$y_n$};
%     \node [latent, above left=of yn] (mu) {$\mu$};
%     \node [latent, above right=of yn] (tau) {$\tau$};

%     \edge {mu, tau} {yn};

%     \tikzstyle{plate caption} = [caption, node distance=0, inner sep=0pt,
%     below=5pt of #1.south]

%     \plate {} {(yn)} {$n=1,\ldots,10$};
%   \end{tikzpicture}
%   \caption{Graphical model of the example problem.}
%   \label{fig:graphical_model}
% \end{figure}
% In order to demonstrate BayesPy, this section solves an extremely simple problem
% but which includes the main steps of using BayesPy.  The task is to estimate the
% unknown mean and precision parameters of a Gaussian distribution given ten
% observations.  Thus, the likelihood function is Gaussian:
% \begin{align*}
%   p(y_1,\ldots,y_{10}|\mu,\tau) &= \prod^{10}_{n=1}\mathcal{N}(y_n | \mu, \tau),
% \end{align*}
% where the Gaussian distribution is parameterized by precision instead of
% variance.  The unknown mean $\mu$ and precision $\tau$ are given broad Gaussian
% and gamma priors, respectively:
% \begin{align*}
%   p(\mu) &= \mathcal{N}(\mu|0, 10^{-6}), & p(\tau) &= \mathcal{G}(\tau|10^{-6},
%   10^{-6}).
% \end{align*}
% Figure~\ref{fig:graphical_model} shows the graphical model of this simple model.
% The model is constructed in BayesPy by creating nodes:
% \begin{lstlisting}
% from bayespy.nodes import GaussianARD, Gamma
% mu = GaussianARD(0, 1e-6)
% tau = Gamma(1e-6, 1e-6)
% \end{lstlisting}
% These two nodes represent the two unknown variables that we are interested in.
% They are the parameters of the Gaussian distribution of the observed variable:
% \begin{lstlisting}
% y = GaussianARD(mu, tau, plates=(10,))
% \end{lstlisting}
% The number of observations is given as plates which define the number of
% repetitions.  Note that the syntax of the model construction is similar to the
% mathematical formulation which defines the conditional distributions for each
% variable.  Thus, it is easy to understand the model from the Python code.  After
% the model has been constructed, some nodes are marked as observed and given
% data:
% \begin{lstlisting}
% y.observe([4.5, 3.9, 6.3, 5.6, 4.9, 2.8, 7.4, 6.1, 4.8, 2.1])
% \end{lstlisting}
% Next, we want to find the posterior approximation for our latent variables.  We
% create the VB inference engine:
% \begin{lstlisting}
% from bayespy.inference import VB
% Q = VB(y, mu, tau)
% \end{lstlisting}
% The inference engine takes all the stochastic nodes of the model as input.  The
% message passing algorithm can be run by calling \texttt{update} method:
% \begin{lstlisting}
% Q.update(repeat=100)
% \end{lstlisting}
% The algorithm is run for 100 iterations or until it converges.  In this case,
% the algorithm converges in four iterations.  The posterior approximation can be
% examined, for instance, by plotting the probability density functions of the
% latent variables:
% \begin{lstlisting}
% import bayespy.plot as bpplt
% import numpy as np
% bpplt.pyplot.figure()
% bpplt.pdf(Q[mu], np.linspace(0, 10, num=300))
% bpplt.pyplot.figure()
% bpplt.pdf(Q[tau], np.linspace(0, 2, num=300))
% bpplt.pyplot.show()
% \end{lstlisting}
% Figure~\ref{fig:posterior} shows the resulting approximate posterior
% distributions.  The module \texttt{bayespy.plot} contains functions for simple
% plotting tasks, such as plotting the probability density function or the Hinton
% diagram of a variable.  Note that the standard \texttt{pyplot} module of
% Matplotlib is available at the \texttt{plot} module of BayesPy.
% \begin{figure}
%   \begin{center}
%     \begin{subfigure}[b]{0.4\textwidth}
%       \includegraphics[width=\textwidth]{fig_mu}
%       \caption{$q(\mu)$}
%     \end{subfigure}
%     \begin{subfigure}[b]{0.4\textwidth}
%       \includegraphics[width=\textwidth]{fig_tau}
%       \caption{$q(\tau)$}
%     \end{subfigure}
%   \end{center}
%   \caption{The posterior approximations of $\mu$ and $\tau$.}
%   \label{fig:posterior}
% \end{figure}
